The Semantics and Implementation of \CEU: a Synchronous Reactive Language based on Esterel

### 2015-09-23 paper.tex
Trabalhos relacionados (incompleto)

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
% Artigo de Rust, nao resolve os problemas
\end{itemize}

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during the entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, however, internal events follow a stack-based execution policy, 
similar to subroutine calls in typical programming 
languages~\cite{ceu.sensys13}, contrasting with queue-based event handling 
adopted for external events.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} can never run concurrently, because each trail segment reacts to a 
different input event and, according to the synchronous model, each reaction is 
atomic.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} occur in the same reaction to input \code{A}.

@@

Regardless of the static analysis, when multiple trails are active at a time in 
\CEU, they are scheduled in the order they appear in the program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as SOL~\cite{wsn.sol} and 
PRET-C\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.

@@

\end{comment}
\section{Implementation}

@@

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The trails translate to a vector of \code{trail\_t} elements with the maximum 
number of simultaneous trails (lines 12--16).
Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
all others are set to \code{INACTIVE}.

@@

@@

- Kiel Esterel compiler

@@

PRET-C
http://embedded.cs.uni-saarland.de/publications/BuildingTimingPredictableEmbeddedSystems.pdf

>>>>

\begin{itemize}
\item Relevancia de Esterel
\item Ceu
    \begin{itemize}
        \item Foco inicial em "constrained WSNs"
        \item Estendendo para outros dominios? (jogos, artigo Mod'15)
        \item Usos: SenSys, Terra, sala de aula, GSoC
        \item ideia de tamanhos, ROM,RAM,CPU
    \end{itemize}
\item Limitacoes do modelo sincrono
\item Semantica e Implementacao (nesse artigo, somente subset estatico)
% Artigo de Rust, nao resolve os problemas
%https://blog.skcript.com/asynchronous-io-in-rust-36b623e7b965
%state machines everywhere
\end{itemize}

%Our work focuses on \emph{concurrency safety}, rather than \emph{type 
%safety}~\cite{wsn.safety}.%
%\footnote{
%We consider both safety aspects to be complimentary and orthogonal, i.e., 
%type-safety techniques could also be applied to \CEU.
%}

@@

Esterel makes no semantic distinctions between internal and external signals, 
both having only the notion of either presence or absence during an entire 
reaction~\cite{esterel.preemption}.
%
In \CEU, external input events are unique within reactions, with
intrinsic queue-based handling, and programs cannot emit them.
%
In contrast, internal events follow a stack-based execution policy, similar to 
subroutine calls in typical programming languages~\cite{ceu.sensys13}.
%
Figure~\ref{lst.prints} illustrates the use of internal signals (events) in 
Esterel and \CEU.
%
In the version in Esterel, on the occurrence of \code{A}, \code{B} is emitted 
and they both become active, resulting in the invocation of \code{f()} and 
\code{g()} in no particular order.
%
In the version in \CEU, the occurrence of \code{A} makes the program behave as 
follows (with the stack contents in italics):
%
{\small
\begin{enumerate}
\setlength{\itemsep}{0pt}
\item 1st trail awakes (line 5), emits \code{b}, and pauses.\\
    \emph{stack: [1st-trail]}
\item 2nd trail awakes (line 9), calls \code{\_f()}, and terminates.\\
    \emph{stack: [1st-trail]}
\item 1st trail (on top of the stack) resumes, calls \code{\_g()}, and 
    terminates.\\
    \emph{stack: []}
\item Both trails have terminated, so the \code{par/and} rejoins, and the 
program also terminates;
\end{enumerate}
}

@@

Concurrency in \CEU is characterized when two or more trail segments in 
parallel execute during the same reaction chain.
A trail segment is a sequence of statements followed by an \code{await} (or 
termination).
%
In the program in the left of Figure~\ref{lst.shared}, the two assignments to 
\code{x} can never run concurrently, because each trail segment reacts to a 
different input event and, according to the semantics of \CEU, cannot occur 
simultaneously.
%
However, the program in the right is non-deterministic, because the two 
assignments to \code{y} occur in the same reaction to input \code{A}.

@@

Regardless of the static analysis, when multiple trails are active at a time in 
\CEU, they are scheduled in the order they appear in the program source code.
%
Therefore, even though the program in the right of Figure~\ref{lst.shared} is 
suspicious, the assignments to \code{y} both atomic and deterministic, i.e., 
after \code{A} occurs, \code{y} is $4$ (\code{(1+1)*2}).
%
On the one hand, enforcing an execution order for concurrent operations may 
seen arbitrary and also precludes true parallelism.
On the other hand, it provides a priority scheme for trails, and makes 
shared-memory concurrency more tractable.
%
For constrained embedded development, we believe that deterministic 
shared-memory concurrency is beneficial, given the extensive use of memory 
mapped ports for I/O and the lack of hardware support for real parallelism.
Other synchronous embedded languages, such as \emph{SOL}~\cite{wsn.sol} and 
\emph{PRET-C}\cite{rp.pretc}, made a similar design choice.
%For Esterel, however, is also used in hardware design~\cite{rp.twelve} where 
%parallelism is inherent.

Figure~\ref{lst.det} compares the two syntactically equivalent code fragments 
in Esterel and \CEU to summarize the semantic difference regarding 
(non-)determinism.
%
Even though the program in \CEU executes deterministically, the compiler still 
issues a warning, because an apparently innocuous reordering of trails modifies 
the semantics of the program.
%
Note that in Esterel multiple external events can coexist in the same reaction, 
which disallows a similar temporal analysis.

@@

\end{comment}

\section{Implementation}

@@

In the final generated code in \emph{C}, each trail label representing an entry 
point becomes a \emph{switch case} with the associated code to execute.
%
Figure~\ref{lst.impl.trails} illustrates the generation process.
For the program in the left of the figure, the compiler extracts the entry 
points and associated trails, e.g., the label \code{Awake\_e} will execute on 
\code{TRAIL-0} (line 7).
%
For each statement that yields (\code{emit} and \code{await}), resumes 
(\code{par/and}, \code{par/or}, and \code{finalize}), or aborts (\code{par/or} 
and \code{break}), the compiler splits the trail into segments with associated 
entry points.
%
The entry points translate to an \code{enum} in the generated code (lines 
1--10, in the middle of the figure).
The trails translate to a vector of \code{trail\_t} elements with the maximum 
number of simultaneous trails (lines 12--16).
Initially, \code{TRAIL-0} is set to execute the \code{Main} entry point, while 
all others are set to \code{INACTIVE}.

@@

\section{Terra and the \CEU-VM}

@@

\emph{(As subsecoes serao retiradas do texto ao final...)}

\subsection{Semantics}

\CEU has a strong influence from Esterel~\cite{esterel.ieee91} and embraces the 
disciplined synchronous-reactive model with support for lexical composition of 
lines of execution.
%
However, there are fundamental semantic differences that prevents the design of 
\CEU as pure extensions to Esterel (i.e., on top of its formal semantics).
%
In particular, Esterel has a notion of time similar to that of digital circuits 
in which multiple signals can be active at a clock tick (Esterel is also used 
in hardware design).
In \CEU, instead of clock ticks, the occurrence of a single external event that 
defines a time unit.
We believe that for software design, this approach simplifies the reasoning 
about concurrency.
For instance, the uniqueness of external events in \CEU is a prerequisite for 
the temporal analysis that enables safe shared-memory concurrency.
%
Furthermore, \CEU also distinguishes external events from stack-based internal 
events, which provides a limited form of subroutines that supports \CEU's 
reactive statements.

A number of synchronous languages have been designed to interoperate with C, 
such as \emph{Reactive C}~\cite{rp.rc}, Protothreads \cite{wsn.protothreads}, 
\emph{PRET-C}~\cite{rp.pretc} and \emph{SC}~\cite{rp.synchc}.
%
They offer Esterel-like parallel compositions with communication via shared 
variables, relying on deterministic semantics to preserve determinism.
%
However, it is the responsibility of the programmer to specify the execution 
order for threads, based on either explicit priorities or source code lexical 
order (like \CEU).
%
These languages have a tick-based notion of time similar to Esterel, also 
disallowing the event-based temporal analysis of \CEU.

\emph{URBI}~\cite{rp.urbi} is a reactive scripting language with a rich set of 
control constructs for time management, event-driven communication, and 
concurrency.
Concurrency is based on stackful coroutines, diverging from our goals regarding 
resource efficiency and static bounds for memory and execution time.

\begin{comment}
More recently, Wireless Sensor Networks (WSNs) emerged as an active research 
area for highly constrained embedded concurrency, resulting in the development 
of many synchronous languages~\cite{wsn.protothreads,wsn.sol,wsn.osm}.
%
offer lightweight cooperative multi-threading for embedded systems.
Its stackless implementation reduces memory consumption but precludes support 
for local variables.
\CEU also avoids the use of stacks for trails, but preserves support for locals 
by calculating the required memory at compile time.
%
SOL~\cite{wsn.sol} and OSM~\cite{wsn.osm} provide parallel state machines for 
WSNs, offering a formal and mature model for programming embedded systems.
However, the main contributions of \CEU, stacked execution for internal 
events and safe support for shared-memory concurrency, do not directly adapt to 
the state-machine formalism.

In common among the referred works is the agreement in providing low-level 
access (e.g., systems calls and shared-memory) and lock-free concurrency that 
precludes race conditions on programs.
However, they do not specify an execution order for tasks reacting to the same 
external stimulus~\cite{esterel.primer,wsn.protothreads,wsn.osm}.
This way, if two tasks access the same resource concurrently, even if the 
accesses are race free, the final outcome is nondeterministic.
As discussed in Section~\ref{sec.safety.det}, \CEU refuses programs with such 
behavior.

On the opposite side of the spectrum of concurrency models, asynchronous 
languages for embedded systems~\cite{wsn.mantisos,arduino.occam}
assume time independence among processes and are more appropriate for 
applications with a low synchronization rate or for those involving
algorithmic-intensive problems.
\end{comment}

\subsection{Implementation}

Esterel has different compilation backends that synthesizes to software and 
also to hardware circuits~\cite{esterel.emperor,esterel.tutorial}.
Among the software-based approaches, \emph{SAXO--RT}~\cite{esterel.saxort} is 
the closest to our implementation for trail allocation and scheduling:
The compiler slices programs into ``control points'' (analogous to our ``entry 
points'') and rearranges them into a directed acyclic graph respecting the 
constructive semantics of Esterel.
Then, it flattens the graph into sequential code in \emph{C} suitable for 
static scheduling.

\subsection{Virtual Machines}

\begin{comment}
- Kiel Esterel compiler

@@

http://embedded.cs.uni-saarland.de/publications/BuildingTimingPredictableEmbeddedSystems.pdf
survey of timeing-predictable embedded systems,
synchronous languages do some at the lang semantic level
\cite{pret.building}
\end{comment}
